---
title: "Milestone 3"
output:
  html_document: default
  word_document: default
date: "2025-10-12"
---


# Data Dictionary

Below is the data dictionary for the key variables included in the cleaned and combined morbidity dataset.

| Variable name | Data type | Description |
|---------------|-----------|-------------|
| week_start | Date | First day of the reporting week in which cases were recorded. |
| county | Character | County of residence associated with the case counts. |
| age_group | Categorical (character) | Age classification for each demographic stratum (e.g., "0-17", "18-49"). |
| sex | Categorical (character) | Sex of individuals in the demographic stratum. |
| cases_new | Numeric | Number of newly reported infections in the reporting week. |
| cases_cum | Numeric | Total cumulative infections reported up to that week. |




```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE, warning=FALSE, message=FALSE}
# load packages
library(readr)
library(dplyr)
library(tidyverse)
library(janitor)
library(lubridate)
library(readr)
library(knitr)

```


```{r}

# 1) california population 
ca_pop <- read_csv("ca_pop_2023.csv", show_col_types = FALSE)

# 2) statewide dataset
ca_state <- read_csv(
  "sim_novelid_CA.csv",
  na = c("", "NA", "na"),
  show_col_types = FALSE
)

# 3) los angeles county dataset
la_county <- read_csv(
  "sim_novelid_LACounty.csv",
  na = c("", "NA", "na"),
  show_col_types = FALSE
)
```




```{r}
#la county has column DT_REPORT that doesn't have any meaningful data (ONLY NA)

la_county <- la_county %>% 
  select(-DT_REPORT)

#ca_pop has race7 column name so I changed it to race
ca_pop<-ca_pop %>% rename(race=race7)

#ca_state has category race_ethnicity as numeric, but codes should be categorical
ca_state <- ca_state %>%
  mutate(race_ethnicity =
           as.factor(race_ethnicity))


```


Clean up column names (recommend using snake case, which is lower case text with underscores between words)

```{r}

#Changed la county data columns to lower case
names(la_county) <- tolower(names(la_county))

#ca_pop has race7 column name so I changed it to race in code chunk above.

#The data columns in ca_state data are already in snake case
```



```{r}

# 1) Clean names (snake_case, strip weird chars)
ca_state  <- ca_state  %>% janitor::clean_names()
la_county <- la_county %>% janitor::clean_names()

# 2) See what's mismatched (diagnostics)
setdiff(names(ca_state),  names(la_county))  # in CA not in LA
setdiff(names(la_county), names(ca_state))   # in LA not in CA

#3) Harmonize to a common schema
schema_map <- list(
  week_start      = c("week_start","week_of","weekstart","week"),
  county          = c("county","county_name"),
  cases_new       = c("cases_new","new_cases","incident_cases","new"),
  cases_cum       = c("cases_cum","cumulative_cases","cum_cases","cases_total"),
  severity        = c("severity","case_severity","severity_level"),
  recovered       = c("recovered","recovery_status","recovered_flag"),
  age_group       = c("age_group","agegrp","age_cat","age_category"),
  sex             = c("sex","gender"),
  race_ethnicity  = c("race_ethnicity","race_eth","race","ethnicity","race_ethnicity_combined")
)

harmonize_names <- function(df, schema_map){
  out <- df
  for (std in names(schema_map)) {
    hit <- intersect(schema_map[[std]], names(out))
    if (length(hit) == 0) {
      out[[std]] <- NA
    } else if (hit[1] != std) {
      out <- out %>% rename(!!std := all_of(hit[1]))
    }
  }
  out
}

ca_h <- harmonize_names(ca_state,  schema_map)
la_h <- harmonize_names(la_county, schema_map)

# ---- 4) 

# --- Standardize CA statewide ---
ca_std <- ca_h %>%
  mutate(
    week_start = suppressWarnings(lubridate::ymd(dt_diagnosis))   # CA uses dt_diagnosis
  )

# --- Standardize LA County ---
la_std <- la_h %>%
  mutate(
    week_start = suppressWarnings(lubridate::ymd(dt_dx))          # LA uses dt_dx
  )

# ---- 5) Align columns and combine ----
core_cols <- c("week_start","county","cases_new","cases_cum",
               "severity","recovered","age_group","sex","race_ethnicity","source")

ca_h <- ca_std %>% select(any_of(core_cols), everything())
la_h <- la_std %>% select(any_of(core_cols), everything())

combined <- bind_rows(
  ca_std %>% select(any_of(core_cols), everything()),
  la_std %>% select(any_of(core_cols), everything())
)

combined <- bind_rows(ca_h, la_h)

# ---- 6 Quick checks ----
# Make sure week_start is properly converted from text to date
combined <- combined %>%
  mutate(week_start = suppressWarnings(lubridate::ymd(week_start)))

# Remove rows where week_start failed to convert
combined <- combined %>% filter(!is.na(week_start))

# Compute summary info
date_min <- min(combined$week_start, na.rm = TRUE)
date_max <- max(combined$week_start, na.rm = TRUE)

list(
  remaining_mismatch_ca_vs_la = c(
    setdiff(names(ca_std), names(la_std)),
    setdiff(names(la_std), names(ca_std))
  ),
  n_rows     = nrow(combined),
  n_counties = dplyr::n_distinct(combined$county),
  date_min   = date_min,
  date_max   = date_max
)

list(
  remaining_mismatch_ca_vs_la = c(setdiff(names(ca_h), names(la_h)),
                                  setdiff(names(la_h), names(ca_h))),
  n_rows    = nrow(combined),
  n_counties= dplyr::n_distinct(combined$county),
 date_min  = min(combined$week_start, na.rm = TRUE),
  date_max  = max(combined$week_start, na.rm = TRUE)
)

# ---- 7) Save for later milestones
readr::write_csv(combined, "combined_morbidity_CA_2023.csv")



```

```{r}


system("git status")
```



```{r}

# Clean names just once
ca_state  <- ca_state  %>% clean_names()
la_county <- la_county %>% clean_names()

#Standardize CA statewide
ca_std <- ca_state %>%
  rename(
    week_start        = dt_diagnosis,
    age_group         = age_cat,
    cases_new         = new_infections,
    cases_cum         = cumulative_infected,
    unrecovered_new   = new_unrecovered,
    unrecovered_cum   = cumulative_unrecovered,
    severe_new        = new_severe,
    severe_cum        = cumulative_severe,
    race_eth          = race_ethnicity
  ) %>%
  mutate(
    week_start = as.Date(week_start),
    race_eth   = as.character(race_eth),
    county     = as.character(county),
    source     = "CA"
  )

# --- Standardize LA County ---
la_std <- la_county %>%
  rename(
    week_start        = dt_dx,
    age_group         = age_category,
    cases_new         = dx_new,
    cases_cum         = infected_cumulative,
    unrecovered_cum   = unrecovered_cumulative,
    race_eth          = race_eth
  ) %>%
  mutate(
    # dt_dx like "29MAY2023" → parse with dmy()
    week_start = dmy(week_start),
    race_eth   = as.character(race_eth),
    county     = "Los Angeles County",
    source     = "LA"
  ) %>%
  # make sure all columns that exist in CA exist here too
  mutate(
    new_severe      = if (!"severe_new" %in% names(.)) NA_integer_ else severe_new,
    severe_cum      = if (!"severe_cumulative" %in% names(.)) NA_integer_ else severe_cumulative
  )

# Align columns and bind
core_cols <- c(
  "county","week_start","age_group","sex","race_eth",
  "cases_new","cases_cum","unrecovered_new","unrecovered_cum",
  "severe_new","severe_cum","source"
)

# Add any missing columns with NA to each side, then select in same order
add_missing <- function(df, cols){
  missing <- setdiff(cols, names(df))
  if (length(missing)) df[missing] <- NA
  df %>% select(all_of(cols))
}

ca_std2 <- add_missing(ca_std, core_cols)
la_std2 <- add_missing(la_std, core_cols)

combined <- bind_rows(ca_std2, la_std2)

# --- Quick checks ---
list(
  n_rows          = nrow(combined),
  n_counties      = dplyr::n_distinct(combined$county),
  date_min        = min(combined$week_start, na.rm = TRUE),
  date_max        = max(combined$week_start, na.rm = TRUE),
  any_na_dates    = any(is.na(combined$week_start))
)

#test
```


## Descriptive Statistics

### Summary for Numeric Variable: cases_new

```{r}

# Ensure the combined dataset is loaded
combined <- read_csv("combined_morbidity_CA_2023.csv", show_col_types = FALSE)

# Numeric summary for cases_new
numeric_summary <- combined %>%
  filter(!is.na(cases_new)) %>%
  summarise(
    mean_cases_new = mean(cases_new),
    median_cases_new = median(cases_new),
    range_cases_new = paste0(min(cases_new), " – ", max(cases_new))
  )

knitr::kable(numeric_summary, caption = "Summary Statistics for Cases New")
```

---

### Frequency Table: County
```{r}
knitr::kable(table(combined$county), caption = "Frequency of Counties")
```

---

### Frequency Table: Age Group
```{r}
knitr::kable(table(combined$age_group), caption = "Frequency of Age Groups")
```

---

### Frequency Table: Sex
```{r}
knitr::kable(table(combined$sex), caption = "Frequency of Sex")
```


### Frequency Table: Severity

```{r}
knitr::kable(table(combined$severity), caption = "Frequency of Severity Categories")
```

### Frequency Table: Race/Ethnicity

```{r}
knitr::kable(table(combined$race_ethnicity), caption = "Frequency of Race/Ethnicity")
```


