---
title: 'PH251W Group 6 Milestone #2'
output: pdf_document
date: "2025-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Description of dataset:

What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
How does the dataset relate to the group problem statement and question?

sim_novelid_LACounty.csv: 

1. The Los Angeles County dataset is a  file that includes weekly data on new and cumulative cases, severity, and recovery for 2023. It is broken down by age, race/ethnicity, and sex.

2. This dataset attempts to fill in the gap for Los Angeles County, which was not included in the statewide file. It lets us look at whether certain groups in Los Angeles are more affected and helps guide decisions on how resources should be allocated for prevention and treatment.

sim_novalid_CA.csv:

ca_pop_2023.csv:

1. This population dataset estimates data by demographic category and county for 2023.
2. The dataset is will be used for county comparisons to normalize for populations to avoid misleading conclusions. We will also use this daya set to help identify demographics that are disproportionately affected.  

Import statements for all datasets:

Document the import process

```{r}
# load packages
library(readr)
library(dplyr)
library(tidyverse)


# 1) california population 
ca_pop <- read_csv("ca_pop_2023.csv", show_col_types = FALSE)

# 2) statewide dataset
ca_state <- read_csv(
  "sim_novelid_CA.csv",
  na = c("", "NA", "na"),
  show_col_types = FALSE
)

# 3) los angeles county dataset
la_county <- read_csv(
  "sim_novelid_LACounty.csv",
  na = c("", "NA", "na"),
  show_col_types = FALSE
)
```


Use appropriate import function and package based on the type of file
Utilize function arguments to control relevant components (i.e. change column types, column names, missing values, etc.)

```{r}
#la county has column DT_REPORT that doesn't have any meaningful data (ONLY NA)

la_county <- la_county %>% 
  select(-DT_REPORT)

#ca_pop has race7 column name so I changed it to race
ca_pop<-ca_pop %>% rename(race=race7)


```


Clean up column names (recommend using snake case, which is lower case text with underscores between words)

```{r}

#Changed la county data columns to lower case
names(la_county) <- tolower(names(la_county))

#ca_pop has race7 column name so I changed it to race in code chunk above.


```

Identify data types for 5+ data elements/columns/variables:

Identify 5+ data elements required for your specified scenario. If <5 elements are required to complete the analysis, please choose additional variables of interest in the data set to explore in this milestone.


Utilize functions or resources in RStudio to determine the types of each data element (i.e. character, numeric, factor)


Identify the desired type/format for each variable—will you need to convert any columns to numeric or another type? 


Provide a basic description of the 5+ data elements:

Numeric: mean, median, range
Character: unique values/categories
Or any other descriptives that will be useful to the analysis
Describe cleaning that each data source may need:

```{r}


```

