---
title: 'Updated Milestone #4'
output: html_document
date: "2025-11-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

Description of dataset:

What is the data source? (1-2 sentences on where the data is coming from, dates included, etc.)
How does the dataset relate to the group problem statement and question?

sim_novelid_LACounty.csv: 

1. The Los Angeles County dataset is a  file that includes weekly data on new and cumulative cases, severity, and recovery for 2023. It is broken down by age, race/ethnicity, and sex.

2. This dataset attempts to fill in the gap for Los Angeles County, which was not included in the statewide file. It lets us look at whether certain groups in Los Angeles are more affected and helps guide decisions on how resources should be allocated for prevention and treatment.

sim_novelid_CA.csv:

1. The sim_novelid_CA dataset contains aggregated surveillance summary data reported weekly from 5/29/2023 until 12/25/2023 for the novel infectious disease in California. It includes data about disease incidence, disease severity, recovery status, demographic information (age, sex, race/ethnicity) and geographic data (county) that helps to identify when, where, and among which subsets of the population cases are occurring. 

2. The dataset provides information that, in combination with the other datasets, could provide insight  on the course of the outbreak, whether it is disproportionately affecting subsets of the population (demographic or geographic populations), and the populations/counties that would most benefit from resources being allocated. 

ca_pop_2023.csv:

1. This population dataset estimates data by demographic category and county for 2023.

2. The dataset is will be used for county comparisons to normalize for populations to avoid misleading conclusions. We will also use this daya set to help identify demographics that are disproportionately affected.  

Import statements for all datasets:

Document the import process

```{r}
# load packages
library(readr)
library(dplyr)
library(tidyverse)


# 1) california population 
ca_pop <- read_csv("ca_pop_2023.csv", show_col_types = FALSE)

# 2) statewide dataset
ca_state <- read_csv(
  "sim_novelid_CA.csv",
  na = c("", "NA", "na"),
  show_col_types = FALSE
)

# 3) los angeles county dataset
la_county <- read_csv(
  "sim_novelid_LACounty.csv",
  na = c("", "NA", "na"),
  show_col_types = FALSE
)
```


Use appropriate import function and package based on the type of file
Utilize function arguments to control relevant components (i.e. change column types, column names, missing values, etc.)

```{r}
#la county has column DT_REPORT that doesn't have any meaningful data (ONLY NA)

la_county <- la_county %>% 
  select(-DT_REPORT)

#ca_pop has race7 column name so I changed it to race
ca_pop<-ca_pop %>% rename(race=race7)

#ca_state has category race_ethnicity as numeric, but codes should be categorical
ca_state <- ca_state %>%
  mutate(race_ethnicity =
           as.factor(race_ethnicity))


```


Clean up column names (recommend using snake case, which is lower case text with underscores between words)

```{r}

#Changed la county data columns to lower case
names(la_county) <- tolower(names(la_county))

#ca_pop has race7 column name so I changed it to race in code chunk above.

#The data columns in ca_state data are already in snake case
```

Identify data types for 5+ data elements/columns/variables:

Identify 5+ data elements required for your specified scenario. 
If <5 elements are required to complete the analysis, please choose additional variables of interest in the data set to explore in this milestone.

1. dt_dx

2. age_category (Age Group) 

3. sex (Sex of Case) – 

4. race_eth or race_ethnicity 

5. county

6. dx_new or new_infections

7. infected_cumulative or cumulative_infected

8. unrecovered_new or new_unrecovered

9. severe_new or new_severe

10. pop
#I listed 10 just to see if these would be good. Please feel free to edit.

Utilize functions or resources in RStudio to determine the types of each data element (i.e. character, numeric, factor)

```{r}
str(la_county)
str(ca_pop)
str(ca_state)

```



```{r}
library(dplyr)
glimpse(la_county)
glimpse(ca_pop)
glimpse(ca_state)
```

```{r}
library(dplyr)
glimpse(la_county)
glimpse(ca_pop)

```

```{r}
library(dplyr)
glimpse(la_county)
glimpse(ca_pop)
#I used this function too to see what it would give us
```



Identify the desired type/format for each variable—will you need to convert any columns to numeric or another type? 


Provide a basic description of the 5+ data elements:

dt_dx – Date – The calendar date when a case was diagnosed.

age_category – Categorical (string/factor) – Groups cases into age ranges.

sex – Categorical (string/factor) – Identifies the sex of individuals.

race_eth or race_ethnicity  – Categorical (string/factor) – Indicates the racial/ethnic group of the case.

county – Categorical (string/factor) – Identifies the county of residence 

dx_new or new_infections – Numeric (integer) – Number of newly reported infections in a given week.

infected_cumulative or cumulative_infected – Numeric (integer) – Running total of infections to date.

unrecovered_new or new_unrecovered – Numeric (integer) – Number of newly added cases that have not yet recovered.

severe_new or new_severe – Numeric (integer) – Number of cases classified as severe in a given week.

pop – Numeric (integer) – Population denominator by demographic group, from the 2023 estimates dataset.


Numeric: mean, median, range
Character: unique values/categories
Or any other descriptives that will be useful to the analysis
Describe cleaning that each data source may need:

#Feel free to change or delete what you want. I probably missed some. 

ca_pop cleaning:
1) Column names - race7 needed changed to race (see above)
2) Standardize categorical variables - age_cat will need to be changed to an ordered factor so the age groups are in order. 
3) Check for missing values (NA)

ca_state cleaning: 
1) Coded variables need to be converted to categorical factors (and relabeled)
2) Standardize categorical variables such as "MALE" and "FEMALE" to be more consistent and uniform ("Male", "Female", or all lowercase)
3) Check for missing values (NA) 
4) Check for and address duplicates if they exist

la_county cleaning:
1) Remove unnecessary columns - DT_REPORT did not meaningful data (see above)
2) Fix date variable - dt_dx is a character column that will need to be changed to a Date. 
3) Standardize categorical variables - age_category could be converted to an ordered factor, "MALE" and "Female" to be more consistent and uniform ("Male", "Female", or all lowercase)

**Milestone #3 Details**

For Scenario 1: Infectious disease outbreak (simulated) in California

The sim_novelid_CA.csv file and sim_novelid_LACounty.csv file together represent simulated morbidity for the entire state of California. While it’s certainly possible for data from different sources to adhere to a format standard, one of the challenges of working with secondary data is that it often does not work out that way. 

Therefore to prepare each dataset (milestone #3) so that they can be combined into one, whole state, dataset you will need to do the following:

**Morbidity datasets (from CA and LA County)**
Recode column names, values, or formats that are in discordant (dates, etc)
Combine morbidity datasets into a single dataset
Select demographic and geographic strata (s) of interest
Aggregate the data into a new dataframe to  include only one row per strata of interest
Population dataset
Recode values to be consistent with morbidity datasets so they can be joined
Select demographic or geogrphic strata(s) of interest
Create a rate metric and summarize to include only one row per strata of interest

```{r}
library(dplyr)
library(janitor)
library(lubridate)


# 1) Clean names (snake_case, strip weird chars)
ca_state  <- ca_state  %>% janitor::clean_names()
la_county <- la_county %>% janitor::clean_names()

# 2) See what's mismatched (diagnostics)
setdiff(names(ca_state),  names(la_county))  # in CA not in LA
setdiff(names(la_county), names(ca_state))   # in LA not in CA

#3) Harmonize to a common schema
schema_map <- list(
  week_start      = c("week_start","week_of","weekstart","week"),
  county          = c("county","county_name"),
  cases_new       = c("cases_new","new_cases","incident_cases","new"),
  cases_cum       = c("cases_cum","cumulative_cases","cum_cases","cases_total"),
  severity        = c("severity","case_severity","severity_level"),
  recovered       = c("recovered","recovery_status","recovered_flag"),
  age_group       = c("age_group","agegrp","age_cat","age_category"),
  sex             = c("sex","gender"),
  race_ethnicity  = c("race_ethnicity","race_eth","race","ethnicity","race_ethnicity_combined")
)

harmonize_names <- function(df, schema_map){
  out <- df
  for (std in names(schema_map)) {
    hit <- intersect(schema_map[[std]], names(out))
    if (length(hit) == 0) {
      out[[std]] <- NA
    } else if (hit[1] != std) {
      out <- out %>% rename(!!std := all_of(hit[1]))
    }
  }
  out
}

ca_h <- harmonize_names(ca_state,  schema_map)
la_h <- harmonize_names(la_county, schema_map)

# ---- 4) 

# --- Standardize CA statewide ---
ca_std <- ca_h %>%
  mutate(
    week_start = suppressWarnings(lubridate::ymd(dt_diagnosis))   # CA uses dt_diagnosis
  )

# --- Standardize LA County ---
la_std <- la_h %>%
  mutate(
    week_start = suppressWarnings(lubridate::ymd(dt_dx))          # LA uses dt_dx
  )

# ---- 5) Align columns and combine ----
core_cols <- c("week_start","county","cases_new","cases_cum",
               "severity","recovered","age_group","sex","race_ethnicity","source")

ca_h <- ca_std %>% select(any_of(core_cols), everything())
la_h <- la_std %>% select(any_of(core_cols), everything())

combined <- bind_rows(
  ca_std %>% select(any_of(core_cols), everything()),
  la_std %>% select(any_of(core_cols), everything())
)

combined <- bind_rows(ca_h, la_h)

# ---- 6 Quick checks ----
# Make sure week_start is properly converted from text to date
combined <- combined %>%
  mutate(week_start = suppressWarnings(lubridate::ymd(week_start)))

# Remove rows where week_start failed to convert
combined <- combined %>% filter(!is.na(week_start))

# Compute summary info
date_min <- min(combined$week_start, na.rm = TRUE)
date_max <- max(combined$week_start, na.rm = TRUE)

list(
  remaining_mismatch_ca_vs_la = c(
    setdiff(names(ca_std), names(la_std)),
    setdiff(names(la_std), names(ca_std))
  ),
  n_rows     = nrow(combined),
  n_counties = dplyr::n_distinct(combined$county),
  date_min   = date_min,
  date_max   = date_max
)

list(
  remaining_mismatch_ca_vs_la = c(setdiff(names(ca_h), names(la_h)),
                                  setdiff(names(la_h), names(ca_h))),
  n_rows    = nrow(combined),
  n_counties= dplyr::n_distinct(combined$county),
 date_min  = min(combined$week_start, na.rm = TRUE),
  date_max  = max(combined$week_start, na.rm = TRUE)
)

# ---- 7) Save for later milestones
readr::write_csv(combined, "combined_morbidity_CA_2023.csv")



```

```{r}
# Details to fix warning from 315-316 *** 
#head(unique(combined$week_start), 15)

#names(combined)

#head(unique(combined$week_start), 20)

# names(ca_h)

# names(la_h)

# unique(na.omit(ca_h$week_start))[1:15]
# unique(na.omit(la_h$week_start))[1:15]

# names(ca_h)

# names(la_h)

# head(unique(combined$week_start), 20)

#head(unique(ca_h$week_start), 20)
#head(unique(ca_h$dt_dx), 20)

#head(unique(ca_h$dt_diagnosis), 20)

system("git status")
```



```{r}
#library(dplyr)
#library(lubridate)
#library(janitor)

#Clean names just once
ca_state  <- ca_state  %>% clean_names()
la_county <- la_county %>% clean_names()

#Standardize CA statewide
ca_std <- ca_state %>%
  rename(
    week_start        = dt_diagnosis,
    age_group         = age_cat,
    cases_new         = new_infections,
    cases_cum         = cumulative_infected,
    unrecovered_new   = new_unrecovered,
    unrecovered_cum   = cumulative_unrecovered,
    severe_new        = new_severe,
    severe_cum        = cumulative_severe,
    race_eth          = race_ethnicity
  ) %>%
  mutate(
    week_start = as.Date(week_start),
    race_eth   = as.character(race_eth),
    county     = as.character(county),
    source     = "CA"
  )

# --- Standardize LA County ---
la_std <- la_county %>%
  rename(
    week_start        = dt_dx,
    age_group         = age_category,
    cases_new         = dx_new,
    cases_cum         = infected_cumulative,
    unrecovered_cum   = unrecovered_cumulative,
    race_eth          = race_eth
  ) %>%
  mutate(
    # dt_dx like "29MAY2023" → parse with dmy()
    week_start = dmy(week_start),
    race_eth   = as.character(race_eth),
    county     = "Los Angeles County",
    source     = "LA"
  ) %>%
  # make sure all columns that exist in CA exist here too
  mutate(
    new_severe      = if (!"severe_new" %in% names(.)) NA_integer_ else severe_new,
    severe_cum      = if (!"severe_cumulative" %in% names(.)) NA_integer_ else severe_cumulative
  )

# Align columns and bind
core_cols <- c(
  "county","week_start","age_group","sex","race_eth",
  "cases_new","cases_cum","unrecovered_new","unrecovered_cum",
  "severe_new","severe_cum","source"
)

# Add any missing columns with NA to each side, then select in same order
add_missing <- function(df, cols){
  missing <- setdiff(cols, names(df))
  if (length(missing)) df[missing] <- NA
  df %>% select(all_of(cols))
}

ca_std2 <- add_missing(ca_std, core_cols)
la_std2 <- add_missing(la_std, core_cols)

combined <- bind_rows(ca_std2, la_std2)

# --- Quick checks ---
list(
  n_rows          = nrow(combined),
  n_counties      = dplyr::n_distinct(combined$county),
  date_min        = min(combined$week_start, na.rm = TRUE),
  date_max        = max(combined$week_start, na.rm = TRUE),
  any_na_dates    = any(is.na(combined$week_start))
)

#test
```


 Milestone #4 Details:

Please submit an Rmd or Qmd and publish an html file on RPubs with the following:

1. Final datasets for creation of visualization
-Join all datasets together
-Calculate any remaining data elements needed for analysis
-Show code used to create joined dataset, but please do not print full data frame output (showing data structure with str() is okay)

```{r}
## Milestone 4: Create a final joined data set ----

## ------------------------------------------------------------
## Milestone 4: Create a final joined data set
## ------------------------------------------------------------

library(dplyr)
library(ggplot2)
library(scales)

## -------- 1. Standardize MORBIDITY dataset -------------------
combined_std <- combined %>%
  mutate(
    sex = toupper(sex),

    # Convert numeric race codes → labels used in ca_pop
    race = case_when(
      race_eth == "1" ~ "WHITE NON-HISPANIC",
      race_eth == "2" ~ "BLACK NH",
      race_eth == "3" ~ "ASIAN NH",
      race_eth == "4" ~ "MR NH",
      race_eth == "5" ~ "NHPI NH",
      race_eth == "6" ~ "HISPANIC",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-race_eth)   # remove old variable



## -------- 2. Standardize POPULATION dataset ------------------
ca_pop_strata <- ca_pop %>%
  mutate(
    county = paste0(county, " County"),

    # Fix age categories to match morbidity dataset
    age_group = case_when(
      age_cat %in% c("0-4","5-11","12-17") ~ "0-17",
      age_cat == "18-49" ~ "18-49",
      age_cat == "50-64" ~ "50-64",
      age_cat == "65+"   ~ "65+",
      TRUE ~ NA_character_
    ),

    sex = toupper(sex),
    race = toupper(race)
  ) %>%
  group_by(county, age_group, sex, race) %>%
  summarise(population = sum(pop, na.rm = TRUE), .groups = "drop")



## -------- 3. JOIN morbidity + population ---------------------
joined_data_ca <- combined_std %>%
  left_join(
    ca_pop_strata,
    by = c("county", "age_group", "sex", "race")
  ) %>%
  mutate(
    incidence_rate_100k = if_else(
      !is.na(population) & population > 0,
      (cases_new / population) * 100000,
      NA_real_
    )
  )



## -------- 4. Quick checks ------------------------------------
summary(joined_data_ca$population)
summary(joined_data_ca$incidence_rate_100k)

head(
  joined_data_ca %>%
    select(week_start, county, age_group, sex, race,
           cases_new, population, incidence_rate_100k),
  20
)



```

2. Visualizations (at least one per group member)
-One print quality table as requested in scenario
-One print quality plot or chart as requested in scenario
-For groups of 3, one additional print quality table or plot of your choice (can support the requested data in the scenario, or answer a different question using the same data sources)

```{r}
# Print quality table

```



```{r}

# Print quality plot or chart -> Incidence rate per 100,000 by sex in California by sex
joined_data_ca %>%
  group_by(sex) %>%
  summarise(
    total_cases = sum(cases_new, na.rm = TRUE),
    total_population = sum(population, na.rm = TRUE),
    incidence_rate_100k = (total_cases / total_population) * 100000
  )

sex_summary <- joined_data_ca %>%
  group_by(sex) %>%
  summarise(
    total_cases = sum(cases_new, na.rm = TRUE),
    total_population = sum(population, na.rm = TRUE),
    incidence_rate_100k = (total_cases / total_population) * 100000
  )

ggplot(sex_summary,
       aes(x = sex, y = incidence_rate_100k, fill = sex)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = round(incidence_rate_100k, 1)),
            vjust = -0.5, size = 5) +   # <-- LABELS HERE
  scale_fill_manual(values = c("FEMALE" = "coral", "MALE" = "cyan3")) +
  labs(
    title = "Incidence Rate per 100,000 by Sex",
    subtitle = "Simulated Infectious Disease Outbreak, California & LA County",
    x = "Sex",
    y = "Incidence Rate per 100,000"
  ) +
  theme_minimal(base_size = 15)





# Additional


```

Incidence rates were comparable across sex in this population, with females at 9470 cases per 100,000 and males at 9539 cases per 100,000. This indicates no major differences in disease burden based on sex for this outbreak.

```{r}

```




3. For each visual, include
-Code used to generate visual
-Legend (if necessary)
- 1-2 sentence interpretation
- NOTE:
*Please make sure the visual can stand-alone, meaning it includes enough information in title, legend, and footnote so that a person who sees only the visualization could understand what is being presented.
*Please also make sure column names, axis labels, and any other labels are meaningful and not just the name of the variable (ex: "County" rather than "county_name")



4. Html file that is professionally prepared for presentation and published to RPubs
-Only the necessary information is in the output (e.g., suppress entire data frame outputs but showing data structure with str() is okay.
-Code used to import and clean data (milestones 2 and 3) can be included in the rmd/qmd but please use the option include=FALSE in order to prevent the code and any output from. Code chunk would look like {r chunk_name, include=FALSE}
-Utilize warning=FALSE and message=FALSE to suppress any warnings/messages that may be displaying from any chunks
-For this milestone please make sure code for visualizations are included in the final file. Show your work by "echoing" code used in Rmd/Qmd file to create your tables and graphs. ```{r, echo = TRUE} in code chunk preferences.
-Use headers and sub headers to create an organized document
-Make sure you commit and push your rsconnect folder!

